Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-10-29 WSSSPE3/report/wg_appendix_software_credit.tex

During the first day of WSSSPE3 breakout sessions, the group discussed the Entertainment Identifier Registry (EIDR)\footnote{Entertainment Identifier Registry: \url{http://eidr.org}} as a potential model for scientific software.
That system assigns unique Digital Object Identifiers (DOIs)---the same system used for scientific publications---to all content (e.g., movies, television shows) and contributors, along with relevant metadata.
One important use of this system is to track rights and credit for contributors to entertainment works in order to distrubute revenues---similar to the proposed transitive credit concept.

@@

Although far from a standard practice, examples of citing software in 
publications can be found in various scientific communities---notably, 
representative samples can be found in astronomy\footnote{Astronomy software citation examples: \url{https://docs.google.com/document/d/1q9ULl7alA3veL7Qwg7jGteRWeJwlrkvRHSXjvt-rTs0}} and biology~\cite{Howison2015}.
The group recommended collecting similar examples from other communities, then 
developing a software citation principles document in concert with the 
FORCE11 Software Citation Working Group (see \S\ref{SC:plan} for more details), 
following the model of the FORCE11 Data Citation Principles 
document~\cite{DataCitation2014}.

@@

Finally, although a discrete task from software citations themselves, significant 
discussion focused on ensuring software citations are indexed in the same 
manner as publications, allowing the construction of a corresponding 
software citation network.
Currently, software releases can receive citable DOIs via 
Zenodo\footnote{Zenodo, \url{https://zenodo.org}} and 
figshare\footnote{figshare: \url{https://figshare.com}}; however, these 
citations are not processed by indexers such as Web of Science, Scopus, 
or Google Scholar.
Thus, either in parallel or following the primary task, the group will 
need to reach out to these organizations 
Initial conversations with Elsevier\slash Scopus (via Michael Taylor) 
during WSSSPE3 clarified that Scopus is not yet DataCite DOI aware, and 
also does not yet have an internal identifier for software or data (but 
needs\slash plans to add this support).
Taylor said they prefer a ``software article'' with the usual 
article metadata (e.g., authors, citations), and mentioned Zenodo as an 
example -- this proposal seemed to align with group discussions.
Taylor also mentioned another benefit of the software and associated DOI 
on GitHub: in addition to a citation, could obtain statistics on 
usage/downloads/forks, which happens to be what Depsy\footnote{Depsy: \url{https://depsy.org}} is beginning to try to do.

>>>>

During the first day of WSSSPE3 breakout sessions, the group discussed the Entertainment Identifier Registry (EIDR)~\cite{EIDR} as a potential model for scientific software.
That system assigns unique Digital Object Identifiers (DOIs)---the same system used for scientific publications---to all content (e.g., movies, television shows) and contributors, along with relevant metadata.
One important use of this system is to track rights and credit for contributors to entertainment works in order to distrubute revenues---similar to the proposed transitive credit concept.

@@

Although far from a standard practice, examples of citing software in 
publications can be found in various scientific communities---notably, 
representative samples can be found in astronomy~\cite{astronomy_SW_examples} and biology~\cite{Howison2015}.
The group recommended collecting similar examples from other communities, then 
developing a software citation principles document in concert with the 
FORCE11 Software Citation Working Group (see \S\ref{SC:plan} for more details), 
following the model of the FORCE11 Data Citation Principles 
document~\cite{DataCitation2014}.

@@

Finally, although a discrete task from software citations themselves, significant 
discussion focused on ensuring software citations are indexed in the same 
manner as publications, allowing the construction of a corresponding 
software citation network.
Currently, software releases can receive citable DOIs via 
Zenodo~\cite{zenodo-web} and figshare~\cite{figshare-web}; however, these 
citations are not processed by indexers such as Web of Science, Scopus, 
or Google Scholar.
Thus, either in parallel or following the primary task, the group will 
need to reach out to these organizations 
Initial conversations with Elsevier\slash Scopus (via Michael Taylor) 
during WSSSPE3 clarified that Scopus is not yet DataCite DOI aware, and 
also does not yet have an internal identifier for software or data (but 
needs\slash plans to add this support).
Taylor said they prefer a ``software article'' with the usual 
article metadata (e.g., authors, citations), and mentioned Zenodo as an 
example -- this proposal seemed to align with group discussions.
Taylor also mentioned another benefit of the software and associated DOI 
on GitHub: in addition to a citation, could obtain statistics on 
usage/downloads/forks, which happens to be what Depsy\footnote{Depsy: \url{https://depsy.org}} is beginning to try to do.

### 2015-10-29 WSSSPE3/report/wg_main_industry_interaction.tex
fixing a subsection that should have been a subsubsection

\subsection{Topic}

Transition Pathways to Sustainable Software: Industry \& Academic Collaboration

@@

\begin{itemize}
\item A project could be \textbf{refunded}, and development or maintenance of the software continues as planned.
\item A project might located a \textbf{new source of funding} in which case the software may be further developed or simply maintained as before. 
\item The project could transition to a \textbf{community supported model} whereby ownership, maintenance and stewardship of the software becomes similar to peer-production models in open-source. (e.g. see Howison 2015)
\item The project could receive some form of industry sponsorship in which case ownership of the intellectual property, licensing, maintenance activities, hosting, etc. may change significantly. 
\end{itemize}

We characterize each of these potential changes in funding as ``transition pathways'' to sustainable software (see similar work by Geels and Schot, 2007).

At WSSSPE 3 our group was interested in better understanding successful pathways for scientific software to ``transition'' from government-funded research projects to industry sponsorship. (This may be an initially awkward phrase - because some software projects will begin their life being sponsored by industry, or result in collaboration between industry and academia. In such cases, there is still a need to understand how IP and how maintenance of the software is sustained over time)

Although transitions are often studied under the broad umbrella of `technology transfer' we believe there are likely to be a number of different ways in which a pathway from initial production to long-term maintenance and secure funding is achieved. In short, industry sponsorship is an important aspect of sustaining scientific software, but our current understanding of these transitions focuses narrowly on commercial successes / failures.

@@

Examples of the former scenario (benefits accruing to private sector) are as follows:

\begin{itemize}
\item REF Impact Case Studies: \url{http://impact.ref.ac.uk/CaseStudies/}
\item Background of projects funded in the UK: \url{http://gtr.rcuk.ac.uk/}
\item Dowling Review from the UK - addresses complexity of work between these two communities: \url{http://www.raeng.org.uk/policy/dowling-review}
\item Pathway to Impact - UK report  - 2 pages of grant proposals are asked to forecast how they might impact (includes environmental, academic, economic )
\end{itemize}

@@

Our work at WSSSPE3 included the following activities (described in detail below): 1. Brainstorming goals for this type of research, 2. Imagining potential outcomes of completing a set of case studies on this topic, and 3. Generating a set of working definitions for some of the broad concepts we are describing.

@@

We characterize transitions in the following way:

\begin{itemize}
\item Handoff model - i.e., academia initially writes the software, industry (for profit or nonprofit) then takes over project. 
\item Co-Production Model - i.e., industry and academia interact throughout development of the project.
\item Sponsorship Model  - i.e., academia writes and maintains the software, the industry contributes funding for the development/maintenance of software. In this example industry is also likely a user of the software.
\item Spinoff model- transition to for-profit company owned by or in collaboration original developers; or, non-profit formed by or in collaboration with developers. 
\end{itemize}

We characterize sustainability in the following ways:

\begin{itemize}
\item Weak Sustainability: Software continues to be accessible, useful, and usable. 
\item Strong Sustainability: Software meets criteria above, but is also able to be reused for further innovation (i.e., issued non-restrictive open-source license). (see Becker et al., 2014 for extended discussion of weak vs. strong sustainability.)
\end{itemize}

@@

Plans for carrying forward are currently unclear - this project would require sustained attention and effort from our team, and at least some amount of funding in order for us to be involved for extended periods of time.

@@

\subsection{Works Referenced}

\todo{please add these to the bib file and cite them from there.}

\begin{itemize}
\item Becker, C., Chitchyan, R., Duboc, L., Easterbrook, S., Mahaux, M., Penzenstadler, B., ... \& Betz, S. (2014). The Karlskrona manifesto for sustainability design. arXiv preprint arXiv:1410.6968.
\item Geels, F. W., \& Schot, J. (2007). Typology of sociotechnical transition pathways. Research policy, 36(3), 399-417.
\item Howison, J. (2015). Sustaining scientific infrastructures: transitioning from grants to peer production. Proceedings of the 2015 iConference. 
\end{itemize}

>>>>

\subsection{Transition Pathways to Sustainable Software: Industry \& Academic Collaboration}

@@

\begin{itemize}
\item A project could be \textbf{refunded}, and development or maintenance of the software continues as planned.
\item A project might located a \textbf{new source of funding} in which case the software may be further developed or simply maintained as before. 
\item The project could transition to a \textbf{community supported model} whereby ownership, maintenance and stewardship of the software becomes similar to peer-production models in open-source (e.g., see Howison~\cite{howison_sustaining_2015}).
\item The project could receive some form of industry sponsorship in which case ownership of the intellectual property, licensing, maintenance activities, hosting, etc.\ may change significantly. 
\end{itemize}

We characterize each of these potential changes in funding as ``transition pathways'' to sustainable software (see similar work by Geels and Schot~\cite{Geels:2007}).

At WSSSPE 3 our group was interested in better understanding successful pathways for scientific software to ``transition'' from government-funded research projects to industry sponsorship. 
(This may be an initially awkward phrase---some software projects will begin their life being sponsored by industry, or result in collaboration between industry and academia. 
In such cases, there is still a need to understand how IP and how maintenance of the software is sustained over time)

Although transitions are often studied under the broad umbrella of `technology transfer' we believe there are likely to be a number of different ways in which a pathway from initial production to long-term maintenance and secure funding is achieved. In short, industry sponsorship is an important aspect of sustaining scientific software, but our current understanding of these transitions focuses narrowly on commercial successes\slash failures.

@@

Examples of the former scenario (benefits accruing to private sector) are as follows:
\begin{itemize}
\item REF Impact Case Studies: \url{http://impact.ref.ac.uk/CaseStudies/}
\item Background of projects funded in the UK: \url{http://gtr.rcuk.ac.uk/}
\item Dowling Review from the UK: addresses complexity of work between these two communities: \url{http://www.raeng.org.uk/policy/dowling-review}
\item Pathway to Impact - UK report: two pages of grant proposals are asked to forecast how they might impact (includes environmental, academic, economic)
\end{itemize}

@@

Our work at WSSSPE3 included the following activities (described in detail below): (1) brainstorming goals for this type of research, (2) imagining potential outcomes of completing a set of case studies on this topic, and (3) generating a set of working definitions for some of the broad concepts we are describing.

@@

We characterize transitions in the following ways:
\begin{itemize}
\item Handoff model: academia initially writes the software, industry (for profit or nonprofit) then takes over project. 
\item Co-Production Model: industry and academia interact throughout development of the project.
\item Sponsorship Model: academia writes and maintains the software, the industry contributes funding for the development\slash maintenance of software. In this example industry is also likely a user of the software.
\item Spinoff model: transition to for- or non-profit company owned by or in collaboration with original developers. 
\end{itemize}

We characterize sustainability in the following ways:
\begin{itemize}
\item Weak Sustainability: Software continues to be accessible, useful, and usable. 
\item Strong Sustainability: Software meets criteria above, but is also able to be reused for further innovation (i.e., issued non-restrictive open-source license).
\end{itemize}
See Becker et al.~\cite{Becker:2014} for extended discussion of weak vs.\ strong sustainability.

@@

Plans for carrying forward are currently unclear---this project would require sustained attention and effort from our team, and at least some amount of funding in order for us to be involved for extended periods of time.

@@

### 2015-10-29 WSSSPE3/report/wg_main_software_credit.tex

The group discussed a number of topics related to software credit, including a contributorship taxonomy, software citation metadata, standards for citing software in publications, and increasing the value of software in academic promotion and tenure reviews.
Although initial discussions both prior to and during WSSSPE3 focused on contribution taxonomy and dividing credit, discussing as an example the Entertainment Identifier Registry\footnote{Entertainment Identifier Registry: \url{http://eidr.org}} used in the entertainment industry, the group decided to prioritize software citation.
This decision was motivated by the idea that standardizing citations for software would introduce some initial credit for developers, and later the quantification of credit could be refined based on concepts such as transitive credit~\cite{wssspe2_katz,Katz:2014_tc}.

>>>>

The group discussed a number of topics related to software credit, including a contributorship taxonomy, software citation metadata, standards for citing software in publications, and increasing the value of software in academic promotion and tenure reviews.
Although initial discussions both prior to and during WSSSPE3 focused on contribution taxonomy and dividing credit, discussing as an example the Entertainment Identifier Registry~\cite{EIDR} used in the entertainment industry, the group decided to prioritize software citation.
This decision was motivated by the idea that standardizing citations for software would introduce some initial credit for developers, and later the quantification of credit could be refined based on concepts such as transitive credit~\cite{wssspe2_katz,Katz:2014_tc}.
